---
title: "Tidying Data"
author:
  - Luc-Tanton Tran, lqtran@ucdavis.edu
  - William Holley, wjholley@ucdavis.edu
  - Guanzhi Wang, agwang@ucdavis.edu
output: html_document
date: "2025-02-20"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#install.packages("sf")
library(tidyverse)
library(janitor)
library(gtools)
library(sf)
library(ggmap)
```

```{r}
register_stadiamaps("1d92513d-1a7f-407d-8c65-60d0d510c9c0", write = FALSE)
```

## The purpose of this document is to tidy data and rename variables/values for analysis and ease of understanding.

```{r}
tf <- read.csv("data/landscapegenetics/TF_pipits_2011.csv") %>% clean_names() #16 var
ps <- read.csv("data/landscapegenetics/PS_pipits_2016.csv") %>% clean_names() #15 var

tf_1 <- tf %>% select(!autocov1000m) #redefine tf as a different df
tf_ps <- full_join(tf_1,ps) #join all data frames together

# add the word "copies" to all the following
genomics <- read_csv("data/landscapegenetics/TF_PS_pipits.csv") %>%
  rename("5239s1.T_copies" = "5239s1.T",
         "7259s1.A.copies" = "7259s1.A",
         "TLR4_1.G_copies" = "TLR4_1.G",
         "TLR4_2.A_copies" = "TLR4_2.A",
         "TLR4_3.C_copies" = "TLR4_3.C",
         "TLR4_4.C_copies" = "TLR4_4.C") %>% 
  clean_names()
names(genomics) <- sub('^x', "", names(genomics)) #substitute any col names beginning with "x" with a blank string
genomics_tf_ps <- left_join(tf_ps, genomics, by = "sample")
```

#no more scientific notation
```{r}
options(scipen=999)
```

# Rewriting Variables

## Redefine `malaria` as "Y" or "N" (To keep consistient with Plasmodium strains)
```{r}
genomics_tf_ps <- genomics_tf_ps %>% 
  mutate(malaria = if_else(malaria == "1", "Y", "N"))
```

## Age? This one might not be necessary
```{r}
genomics_tf_ps <- genomics_tf_ps %>% 
  mutate(age = if_else(age=="A", "adult", "juvenile")) #I spelled it right this time!
```

_For now, I'll be skipping to the ps/tf df sections because it looks like a pain in the rear to code the heterozygote part. I'll come back when I learn how to make my own function._

## Straight up removing `utm` because what the hell man
```{r}
genomics_tf_ps <- genomics_tf_ps %>% 
  select(!utm)
```
~~Changing `TLR4_het` to homozygous/heterozygous~~ (Found better way)
```{r}
#genomics_tf_ps <- genomics_tf_ps %>% 
#  mutate(tlr4_het = if_else(tlr4_het == 0, "homozygous", "heterozygous"))
```
## Mutating any variable ending with `het` to "homozygous", "heterozygous"
```{r}
genomics_tf_ps <- genomics_tf_ps %>% 
  mutate(across(ends_with("het"), #if the var name ends with "het"
                ~if_else(.x==0, "homozygous", "heterozygous")))
# Check to see if it ends with 0 (homozygous). Otherwise, heterozygous.
# Should we consider shortening it to "homo" and "hetero"? Might be easier to read.
```
## Looking into making some of the dist data into categorical variables...
```{r}
quartileswater <- quantcut(genomics_tf_ps$distwater)
table(quartileswater)

quartilesurban <- quantcut(genomics_tf_ps$dist_urb)
table(quartilesurban)

quartilesfarm <- quantcut(genomics_tf_ps$distfarm)
table(quartilesfarm)

quartilespoul <- quantcut(genomics_tf_ps$distpoul)
table(quartilespoul)
```
```{r}
genomics_tf_ps <- genomics_tf_ps %>% 
  mutate(distwater_cat = case_when(distwater <= 105 ~ "close",
                                   distwater > 105 & distwater <= 258 ~ "median close",
                                   distwater > 258 & distwater <= 725 ~ "median far",
                                   distwater > 725 ~ "far"))
```

```{r}
genomics_tf_ps <- genomics_tf_ps %>% 
  mutate(disturb_cat = case_when(dist_urb <= 45 ~ "close",
                                   dist_urb > 45 & dist_urb <= 108 ~ "median close",
                                   dist_urb > 108 & dist_urb <= 282 ~ "median far",
                                   dist_urb > 282 ~ "far"))
```

```{r}
genomics_tf_ps <- genomics_tf_ps %>% 
  mutate(distfarm_cat = case_when(distfarm <= 429 ~ "close",
                                   distfarm > 429 & distfarm <= 723 ~ "median close",
                                   distfarm > 723 & distfarm <= 1240 ~ "median far",
                                   distfarm > 1240 ~ "far"))
```

```{r}
genomics_tf_ps <- genomics_tf_ps %>% 
  mutate(distpoul_cat = case_when(distpoul <= 860 ~ "close",
                                   distpoul > 860 & distpoul <= 1830 ~ "median close",
                                   distpoul> 1830 & distpoul <= 3410 ~ "median far",
                                   distpoul > 3410 ~ "far"))
```

```{r}
genomics_tf_ps %>% 
  select(sample, distwater_cat, malaria) %>% 
  na.omit() %>% 
  ggplot(aes(x = distwater_cat, y = malaria))+
  geom_col()+
  coord_flip()
```

## we can change the category names at some point but those were the best I could come up with in the moment


## converting "0/1" to "absent/present"
```{r}
genomics_tf_ps <- genomics_tf_ps %>% 
  mutate(across(matches("_[acgt]$|_prot_[1234]$"), ~if_else(.x== 0, "absent", "present")))
```


```{r}
tf_ps_latlong <- st_as_sf(genomics_tf_ps, coords = c("easting", "northing"), crs = "+proj=utm +zone=28 +datum=WGS84 +units=m +north")
tf_ps_latlong <- st_transform(tf_ps_latlong, crs = 4326)
genomics_tf_ps <- genomics_tf_ps %>%
  mutate(longitude = st_coordinates(tf_ps_latlong)[,1],
         latitude = st_coordinates(tf_ps_latlong)[,2])
genomics_tf_ps %>% 
  select(sample, latitude, longitude)
```
```{r}
genomics_tf_ps %>% 
  filter(island == "TF") %>% 
  select(latitude, longitude) %>%
  summary()
```
```{r}
latitude <- c(28.01, 28.58)
longitude <- c(-16.92, -16.18)
tf_bbox <- make_bbox(longitude, latitude, f = 0.05)
tf_map <- get_stadiamap(tf_bbox, maptype = "stamen_terrain", zoom=12)
```

```{r}
ggmap(tf_map)+
  geom_point(data = genomics_tf_ps, aes(longitude, latitude, color = malaria), size=1) +
  labs(x= "Longitude", y= "Latitude", title = "Bird Locations of Tenerife")
```

```{r}
genomics_tf_ps %>%
  filter(island == "PS") %>% 
  select(latitude, longitude) %>% 
  summary()
```

```{r}
latitude <- c(33.03, 33.10)
longitude <- c(-16.39, -16.3)
ps_bbox <- make_bbox(longitude, latitude, f = 0.05)
ps_map <- get_stadiamap(ps_bbox, maptype = "stamen_terrain", zoom=12)
```

```{r}
ggmap(ps_map)+
  geom_point(data = genomics_tf_ps, aes(longitude, latitude, color = malaria), size=1) +
  labs(x= "Longitude", y= "Latitude", title = "Bird Locations of Porto Santo")
```

```{r}
names(genomics_tf_ps)
```



## Update csv file
```{r}
write.csv(genomics_tf_ps, "data/landscapegenetics/genomics_tf_ps.csv", row.names = FALSE)
```


