---
title: "randomanalysis"
author: "gz"
date: "2025-02-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

#load the libraries
```{r message=FALSE, warning=FALSE}
library("tidyverse")
library("janitor")
library("naniar")
library("dplyr")
library("here")#3
library("shiny")
library("shinydashboard")
library("ggplot2")
library("shinythemes")
library("gtools")
library("scales")
```

#load the data
```{r}
tf <- read_csv(here("final project/data/landscapegenetics","TF_pipits_2011.csv")) %>% clean_names() #16 var
ps <- read.csv(here("final project/data/landscapegenetics","PS_pipits_2016.csv")) %>% clean_names() #15 var

tf_1 <- tf %>% select(!autocov1000m) #redefine tf as a different df
tf_ps <- merge(tf_1,ps, all=TRUE) #merge all data frames together
genomics <- read.csv(here("final project/data/landscapegenetics","TF_PS_pipits.csv")) %>% clean_names()
genomics_tf_ps <- read.csv(here("final project/data/landscapegenetics","genomics_tf_ps.csv")) %>% clean_names()
```
#from tlr4_prot_1 to tlr4_prot_4, how they related to malaria infection

```{r}
genomics_tf_ps %>% 
  na.omit %>% 
    
    count(tlr4_prot_1, malaria) %>%
    group_by(tlr4_prot_1) %>%
    mutate(percent = n / sum(n) * 100)%>%
ggplot(aes(x = factor(tlr4_prot_1), y=percent,fill = factor(malaria))) +
    geom_col(position = "dodge") +  
    scale_y_continuous(labels = scales::percent_format(scale = 1)) +  # make ratio from 0~1.0 to 0%~100%
    labs(title = "relationship between genotype and malaria infection ",
         x = "TLR4_prot_1", 
         y = "percent of Infection",
         fill = "malaria_status (N=unaffected, Y=affected)") +
    theme_minimal()
```

#tlr4_prot_2
```{r}
genomics_tf_ps %>% 
  na.omit %>% 
    
    count(tlr4_prot_2, malaria) %>%
    group_by(tlr4_prot_2) %>%
    mutate(percent = n / sum(n) * 100)%>%
ggplot(aes(x = factor(tlr4_prot_2), y=percent,fill = factor(malaria))) +
    geom_col(position = "dodge") +  
    scale_y_continuous(labels = scales::percent_format(scale = 1)) +  # make ratio from 0~1.0 to 0%~100%
    labs(title = "relationship between genotype and malaria infection ",
         x = "TLR4_prot_2", 
         y = "percent of Infection",
         fill = "malaria_status (N=unaffected, Y=affected)") +
    theme_minimal()
```

#tlr4_prot_3
```{r}
genomics_tf_ps %>% 
  na.omit %>% 
    
    count(tlr4_prot_3, malaria) %>%
    group_by(tlr4_prot_3) %>%
    mutate(percent = n / sum(n) * 100)%>%
ggplot(aes(x = factor(tlr4_prot_3), y=percent,fill = factor(malaria))) +
    geom_col(position = "dodge") +  
    scale_y_continuous(labels = scales::percent_format(scale = 1)) +  # make ratio from 0~1.0 to 0%~100%
    labs(title = "relationship between genotype and malaria infection ",
         x = "TLR4_prot_3", 
         y = "percent of Infection",
         fill = "malaria_status (N=unaffected, Y=affected)") +
    theme_minimal()
```

#tlr4_prot_4
```{r}
genomics_tf_ps %>% 
  na.omit %>% 
    
    count(tlr4_prot_4, malaria) %>%
    group_by(tlr4_prot_4) %>%
    mutate(percent = n / sum(n) * 100)%>%
ggplot(aes(x = factor(tlr4_prot_4), y=percent,fill = factor(malaria))) +
    geom_col(position = "dodge") +  
    scale_y_continuous(labels = scales::percent_format(scale = 1)) +  # make ratio from 0~1.0 to 0%~100%
    labs(title = "relationship between genotype and malaria infection ",
         x = "TLR4_prot_4", 
         y = "percent of Infection",
         fill = "malaria_status (N=unaffected, Y=affected)") +
    theme_minimal()

```


#x7259s1 to malaria
```{r}

genomics_tf_ps %>% 
  na.omit %>% 
   
    count(x7259s1, malaria) %>%
    group_by(x7259s1) %>%
    mutate(percent = n / sum(n) * 100)%>%
ggplot(aes(x = factor(x7259s1), y=percent,fill = factor(malaria))) +
    geom_col(position = "dodge") +  # 
    scale_y_continuous(labels = scales::percent_format(scale = 1)) +  # make ratio from 0~1.0 to 0%~100%
    labs(title = "relationship between genotype and malaria infection ",
         x = "x7259s1", 
         y = "percent of Infection",
         fill = "malaria_status (N=unaffected, Y=affected)") +
    theme_minimal()
```

# how x5239s1 related to malaria, we found out that T allele is key factor
```{r}
genomics_tf_ps %>% 
  na.omit %>% 
   
    count(x5239s1, malaria) %>%
    group_by(x5239s1) %>%
    mutate(percent = n / sum(n) * 100)%>%
ggplot(aes(x = factor(x5239s1), y=percent,fill = factor(malaria))) +
    geom_col(position = "dodge") +  # 
    scale_y_continuous(labels = scales::percent_format(scale = 1)) +  # make ratio from 0~1.0 to 0%~100%
    labs(title = "relationship between genotype and malaria infection ",
         x = "x5239s1", 
         y = "percent of Infection",
         fill = "malaria_status (N=unaffected, Y=affected)") +
    theme_minimal()
```


#according to paper.
allele T is additive affect the malaria infection

```{r}
genomics_tf_ps %>% 
  na.omit %>% 
   filter(island == "TF") %>% 
    count(x5239s1_t, malaria) %>%
    group_by(x5239s1_t) %>%
    mutate(percent = n / sum(n) * 100)%>%
ggplot(aes(x = factor(x5239s1_t), y=percent,fill = factor(malaria))) +
    geom_col(position = "dodge") +  
    scale_y_continuous(labels = scales::percent_format(scale = 1)) +  # make ratio from 0~1.0 to 0%~100%
    labs(title = "relationship between genotype and malaria infection ",
         x = "SNP_5239s1_t", 
         y = "percent of Infection",
         fill = "malaria_status (N=unaffected, Y=affected)") +
    theme_minimal()

genomics_tf_ps %>% 
  na.omit %>% 
   filter(island == "PS") %>% 
    count(x5239s1_t, malaria) %>%
    group_by(x5239s1_t) %>%
    mutate(percent = n / sum(n) * 100)%>%
ggplot(aes(x = factor(x5239s1_t), y=percent,fill = factor(malaria))) +
    geom_col(position = "dodge") +  # 
    scale_y_continuous(labels = scales::percent_format(scale = 1)) +  # make ratio from 0~1.0 to 0%~100%
    labs(title = "relationship between genotype and malaria infection ",
         x = "SNP_5239s1_t", 
         y = "percent of Infection",
         fill = "malaria_status (N=unaffected, Y=affected)") +
    theme_minimal()
    
```

#tlr4_prot_1, haplotype of a immune protein,correlate to low risk of malaria infection

```{r}
genomics_tf_ps %>% 
  na.omit %>% 
   filter(island == "TF") %>% 
    count(tlr4_prot_1, malaria) %>%
    group_by(tlr4_prot_1) %>%
    mutate(percent = n / sum(n) * 100)%>%
ggplot(aes(x = factor(tlr4_prot_1), y=percent,fill = factor(malaria))) +
    geom_col(position = "dodge") +  
    scale_y_continuous(labels = scales::percent_format(scale = 1)) +  # make ratio from 0~1.0 to 0%~100%
    labs(title = "relationship between genotype and malaria infection ",
         x = "TLR4_pro_1", 
         y = "percent of Infection",
         fill = "malaria_status (N=unaffected, Y=affected)") +
    theme_minimal()

genomics_tf_ps %>% 
  na.omit %>% 
   filter(island == "PS") %>% 
    count(tlr4_prot_1, malaria) %>%
    group_by(tlr4_prot_1) %>%
    mutate(percent = n / sum(n) * 100)%>%
ggplot(aes(x = factor(tlr4_prot_1), y=percent,fill = factor(malaria))) +
    geom_col(position = "dodge") +  # 
    scale_y_continuous(labels = scales::percent_format(scale = 1)) +  # make ratio from 0~1.0 to 0%~100%
    labs(title = "relationship between genotype and malaria infection ",
         x = "TLR4_pro_1", 
         y = "percent of Infection",
         fill = "malaria_status (N=unaffected, Y=affected)") +
    theme_minimal()
```


```{r}
genomics_tf_ps %>% 
  filter(island == "TF" & malaria == "Y")%>%  # Select only Tf island
  na.omit() %>%  # Remove NA values 
  group_by(distwater_cat) %>% 
  summarise(n = n()) %>% 
  ggplot(aes(x = reorder(distwater_cat,n),y = n ,fill = factor(distwater_cat)))+ 
  geom_col() +  # Dodge bar chart for comparison 
  labs(title = "TF: Water Distance vs. Malaria Infection",
       x = "Distance to Water",
       y = "Percentage of Infection",
       fill = "Malaria Status (N = Unaffected, Y = Infected)") +
  theme_minimal()
genomics_tf_ps %>% 
  filter(island == "PS" & malaria == "Y") %>%  # 
  na.omit() %>%   
  group_by(distwater_cat) %>% 
  summarise(n = n()) %>%  # 
  ggplot(aes(x = reorder(distwater_cat, n), y = n, fill = factor(distwater_cat))) + 
  geom_col() +  
  labs(title = "PS: Water Distance vs. Malaria Infection",
       x = "Distance to Water",
       y = "Number of Infected Cases",
       fill = "Distance Category") +
  theme_minimal()
```

#how altitude affect
```{r}
alt <- quantcut(genomics_tf_ps$altitude)
table(alt)
altitude1 <- genomics_tf_ps %>% 
mutate(altitude_category = case_when(  
    altitude <= 83 ~ "Low",
    altitude > 83 & altitude <= 169 ~ "Medium",
    altitude > 169 & altitude <= 351 ~ "Medium High",
    altitude > 351 ~ "High" ))
```
mutate(distwater_cat = case_when(distwater <= 105 ~ "close",
                                   distwater > 105 & distwater <= 258 ~ "median close",
                                   distwater > 258 & distwater <= 725 ~ "median far",
                                   distwater > 725 ~ "far"))
```{r}

 altitude1 %>%
  filter( malaria == "Y") %>%  
  na.omit() %>%  
 
  group_by(altitude_category) %>%  
  summarise(n = n()) %>%  
  ggplot(aes(x = reorder(altitude_category,n), y = n, fill = altitude_category)) + 
  geom_bar(stat = "identity") +  
  labs(title = "Tenerife: Malaria Infection by Altitude Category",
       x = "Altitude Category",
       y = "Number of Infected Cases") +
  theme_minimal()
```

#this shinyapp contains TLR4 prot-1 to prot-4 and also x5239s1, and its T allele and x7259s1 gene
```{r}

ui <- fluidPage(
    theme = shinytheme("cosmo"),
    
    titlePanel("TLR4 Protein Variants & Malaria Infection"),
    
    sidebarLayout(
        sidebarPanel(
            selectInput("protein", "Select TLR4 haplotype:", 
                        choices = c("None", "tlr4_prot_1", "tlr4_prot_2", "tlr4_prot_3", "tlr4_prot_4"), 
                        selected = "None"),
            
            selectInput("snp", "Select SNP Genotype:", 
                        choices = c("None", "x5239s1", "x7259s1"), 
                        selected = "None")
        ),
        
        mainPanel(
            # show TLR4 haplotype plots in TF and PS 
            conditionalPanel(
                condition = "input.protein != 'None'",
                h3("TLR4 Protein Variant - TF Island"),
                plotOutput("proteinPlot_TF"),
                h3("TLR4 Protein Variant - PS Island"),
                plotOutput("proteinPlot_PS")
            ),
            
            # SNP part
            conditionalPanel(
                condition = "input.snp == 'x5239s1'",
                h3("SNP_5239s1 - General"),
                plotOutput("plot_x5239s1"),
                h3("SNP_5239s1_T - TF Island"),
                plotOutput("plot_x5239s1_t_TF"),
                h3("SNP_5239s1_T - PS Island"),
                plotOutput("plot_x5239s1_t_PS")
            ),
            
            conditionalPanel(
                condition = "input.snp == 'x7259s1'",
                h3("SNP_7259s1 - General"),
                plotOutput("plot_x7259s1")
            )
        )
    )
)

# Define Server
server <- function(input, output) {
    
    # Plot for TLR4 haplotype (TF Island)
    output$proteinPlot_TF <- renderPlot({
        req(input$protein != "None")
        selected_var <- input$protein
        
        genomics_tf_ps %>%
            na.omit() %>%
            filter(island == "TF") %>%
            count(!!sym(selected_var), malaria) %>%
            group_by(!!sym(selected_var)) %>%
            mutate(percent = n / sum(n) * 100) %>%
            ggplot(aes(x = factor(!!sym(selected_var)), y = percent, fill = factor(malaria))) +
            geom_col(position = "dodge") +
            scale_y_continuous(labels = scales::percent_format(scale = 1)) +
            labs(
                title = paste("Effect of", selected_var, "on Malaria Infection (TF Island)"),
                x = selected_var,
                y = "Percent of Infection",
                fill = "Malaria Status (N=unaffected, Y=affected)"
            ) +
            theme_minimal()
    })

    # Plot for TLR4 haplotype (PS Island)
    output$proteinPlot_PS <- renderPlot({
        req(input$protein != "None")
        selected_var <- input$protein
        
        genomics_tf_ps %>%
            na.omit() %>%
            filter(island == "PS") %>%
            count(!!sym(selected_var), malaria) %>%
            group_by(!!sym(selected_var)) %>%
            mutate(percent = n / sum(n) * 100) %>%
            ggplot(aes(x = factor(!!sym(selected_var)), y = percent, fill = factor(malaria))) +
            geom_col(position = "dodge") +
            scale_y_continuous(labels = scales::percent_format(scale = 1)) +
            labs(
                title = paste("Effect of", selected_var, "on Malaria Infection (PS Island)"),
                x = selected_var,
                y = "Percent of Infection",
                fill = "Malaria Status (N=unaffected, Y=affected)"
            ) +
            theme_minimal()
    })
    
    # Plot for x5239s1
    output$plot_x5239s1 <- renderPlot({
        req(input$snp == "x5239s1")
        
        genomics_tf_ps %>%
            na.omit() %>%
            count(x5239s1, malaria) %>%
            group_by(x5239s1) %>%
            mutate(percent = n / sum(n) * 100) %>%
            ggplot(aes(x = factor(x5239s1), y = percent, fill = factor(malaria))) +
            geom_col(position = "dodge") +
            scale_y_continuous(labels = scales::percent_format(scale = 1)) +
            labs(
                title = "Relationship between x5239s1 Genotype and Malaria Infection",
                x = "x5239s1",
                y = "Percent of Infection",
                fill = "Malaria Status (N=unaffected, Y=affected)"
            ) +
            theme_minimal()
    })
    
    # Plot for x5239s1_t (TF Island)
    output$plot_x5239s1_t_TF <- renderPlot({
        req(input$snp == "x5239s1")
        
        genomics_tf_ps %>%
            na.omit() %>%
            filter(island == "TF") %>% 
            count(x5239s1_t, malaria) %>%
            group_by(x5239s1_t) %>%
            mutate(percent = n / sum(n) * 100) %>%
            ggplot(aes(x = factor(x5239s1_t), y = percent, fill = factor(malaria))) +
            geom_col(position = "dodge") +
            scale_y_continuous(labels = scales::percent_format(scale = 1)) +
            labs(
                title = "Relationship between x5239s1_t Genotype and Malaria Infection (TF Island)",
                x = "SNP_5239s1_t",
                y = "Percent of Infection",
                fill = "Malaria Status (N=unaffected, Y=affected)"
            ) +
            theme_minimal()
    })
    
    # Plot for x5239s1_t (PS Island)
    output$plot_x5239s1_t_PS <- renderPlot({
        req(input$snp == "x5239s1")
        
        genomics_tf_ps %>%
            na.omit() %>%
            filter(island == "PS") %>% 
            count(x5239s1_t, malaria) %>%
            group_by(x5239s1_t) %>%
            mutate(percent = n / sum(n) * 100) %>%
            ggplot(aes(x = factor(x5239s1_t), y = percent, fill = factor(malaria))) +
            geom_col(position = "dodge") +
            scale_y_continuous(labels = scales::percent_format(scale = 1)) +
            labs(
                title = "Relationship between x5239s1_t Genotype and Malaria Infection (PS Island)",
                x = "SNP_5239s1_t",
                y = "Percent of Infection",
                fill = "Malaria Status (N=unaffected, Y=affected)"
            ) +
            theme_minimal()
    })

    # Plot for x7259s1
    output$plot_x7259s1 <- renderPlot({
        req(input$snp == "x7259s1")
        
        genomics_tf_ps %>%
            na.omit() %>%
            count(x7259s1, malaria) %>%
            group_by(x7259s1) %>%
            mutate(percent = n / sum(n) * 100) %>%
            ggplot(aes(x = factor(x7259s1), y = percent, fill = factor(malaria))) +
            geom_col(position = "dodge") +
            scale_y_continuous(labels = scales::percent_format(scale = 1)) +
            labs(
                title = "Relationship between x7259s1 Genotype and Malaria Infection",
                x = "x7259s1",
                y = "Percent of Infection",
                fill = "Malaria Status (N=unaffected, Y=affected)"
            ) +
            theme_minimal()
    })
}

# Run the App
shinyApp(ui = ui, server = server)

```


