---
title: "Midterm 1 W25"
author: "William Holley"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

## Instructions
Answer the following questions and complete the exercises in RMarkdown. Please embed all of your code and push your final work to your repository. Your code must be organized, clean, and run free from errors. Remember, you must remove the `#` for any included code chunks to run. Be sure to add your name to the author header above. 

Your code must knit in order to be considered. If you are stuck and cannot answer a question, then comment out your code and knit the document. You may use your notes, labs, and homework to help you complete this exam. Do not use any other resources- including AI assistance or other students' work.  

Don't forget to answer any questions that are asked in the prompt! Each question must be coded; it cannot be answered by a sort in a spreadsheet or a written response.  

Be sure to push your completed midterm to your repository and upload the document to Gradescope. This exam is worth 30 points. 

Please load the following libraries.
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
```

Disable scientific notation.
```{r message=FALSE, warning=FALSE}
options(scipen=999)
```

In the midterm 1 folder there is a second folder called `data`. Inside the `data` folder, there is a .csv file called `ecs21351-sup-0003-SupplementS1.csv`. These data are from Soykan, C. U., J. Sauer, J. G. Schuetz, G. S. LeBaron, K. Dale, and G. M. Langham. 2016. Population trends for North American winter birds based on hierarchical models. Ecosphere 7(5):e01351. 10.1002/ecs2.1351. This study uses the CBC (Christmas Bird Count) data to estimate population trends for North American winter birds.  

Please load these data as a new object called `ecosphere`. In this step, I am providing the code to load the data, clean the variable names, and remove a footer that the authors used as part of the original publication.
```{r message=FALSE, warning=FALSE}
ecosphere <- read_csv("data/ecs21351-sup-0003-SupplementS1.csv", skip=2) %>%
  #load the data and skip the first two rows
  clean_names() %>%
  #clean the variable names
  slice(1:(n() - 18)) 
  #remove the footer
```

## Questions
Problem 1. (1 point) What are the variable names?

```{r}
names(ecosphere)
```
Problem 2. (1 point) Use the function of your choice to provide a data summary.

```{r}
glimpse(ecosphere)
```


Problem 3. (2 points) How many distinct orders of birds are represented in the data?

```{r}
ecosphere %>% 
  summarize(n_order_birds = n_distinct(order))
```
There are 19 distinct orders of birds represented in the data.


Problem 4. (2 points) Which habitat has the greatest species diversity?

```{r}
ecosphere %>% 
  group_by(habitat) %>% 
  summarize(habitat_n_species_birds = n_distinct(scientific_name)) %>% 
  arrange(-habitat_n_species_birds)
```
The woodland has the greatest species diversity with 177 distinct species represented.


Problem 5. (2 points) For species associated with urban environments, what is the min, max, and mean winter range area?

```{r}
ecosphere %>% 
  filter(urban_affiliate == "Yes") %>% 
  summarize(across(contains("winter"), list(mean = \(x) mean(x, na.rm = T), max = \(x) max(x, na.rm = T), min = \(x) min(x, na.rm = T))))
```


Problem 6. (2 points) As part of our analysis, we need `mass_g` as a new variable. Please convert `log10_mass` to mass in grams (hint: mass_g=10^log10_mass) and store the output as part of the `ecosphere` data.

```{r}
ecosphere <- ecosphere %>% 
  mutate(mass_g = 10^log10_mass)
```


Problem 7. (4 points) Which migratory strategy has the highest average mass (mass_g)?  

```{r}
ecosphere %>% 
  group_by(migratory_strategy) %>% 
  summarize(average_mass_g = mean(mass_g, na.rm = T)) %>% 
  arrange(-average_mass_g)
```


Problem 8. (4 points) Irruptive migratory behavior is characterized by unpredictable movements in response to food availability. What is the average population size for species with irruptive migratory behavior, grouped by habitat and diet?

```{r}
ecosphere %>% 
  filter(migratory_strategy == "Irruptive") %>% 
  group_by(habitat, diet) %>% 
  summarize(irruptive_mean_population_size = mean(population_size, na.rm = T)) %>% 
  arrange(-irruptive_mean_population_size)
```


Problem 9. (4 points). Diet, life expectancy, urban_affiliate, and migratory_strategy are all variables associated with extinction risk or population decline. Which species have a combination of vertebrate diet, long life expectancy, no urban affiliation, and are long-distance migrants? Assuming that the bird with the highest mass is the most at risk, which is the species of greatest concern?  
```{r}
ecosphere %>% 
  filter(diet == "Vertebrates" & life_expectancy == "Long" & urban_affiliate == "No" & migratory_strategy == "Long") %>%
  select(order, family, scientific_name, diet, life_expectancy, urban_affiliate, migratory_strategy, mass_g) %>% 
  arrange(-mass_g)
```
The Phoebastria nigripes is at most risk out of these three birds.


Problem 10. (4 points). Make a new column `conservation_status` that labels species with a population size less than 300,000 as "threatened" and species with a population size greater than 300,000 as "stable". Make sure your results are sorted in descending order. How many species are threatened vs. stable? Based on the results, do you see a problem with this analysis?  

```{r}
ecosphere <- ecosphere %>%
  mutate(conservation_status = ifelse(300000 > population_size, "threatened", "stable"))
ecosphere %>% 
  select(order, family, scientific_name, population_size, conservation_status) %>% 
  arrange(-population_size)
```
```{r}
ecosphere %>% 
  count(conservation_status)
```

There are 250 stable, 28 threatened, and 273 NAs. Almost half of the species of birds here have an entry of NA for population_size, making it difficult to draw conclusions due to incomplete data.



Problem 11. (4 points) Use the `ecosphere` data to perform one exploratory analysis of your choice. The analysis must have a minimum of three lines and two functions. You must also clearly state the question you are attempting to answer.

```{r}
ecosphere %>%
  select(order, family, scientific_name, mean_age_at_sexual_maturity, mean_eggs_per_clutch) %>% 
  mutate(eggs_maturity_ratio = mean_eggs_per_clutch / mean_age_at_sexual_maturity) %>% 
  group_by(order) %>% 
  summarize(mean_eggs_maturity_ratio = mean(eggs_maturity_ratio, na.rm = T)) %>% 
  arrange(-mean_eggs_maturity_ratio)
```
This code creates a new column called eggs_maturity_ratio that explores the ratio of mean_eggs_per_clutch and mean_age_at_sexual_maturity. I then used group_by and summarize to find the average eggs_maturity_ratio for each order of birds, and then used arange to display them in descending order. I was attempting to answer the question of which order of birds has the highest ratio of mean eggs per clutch to mean age at sexual maturity.


## Submit the Midterm
1. Save your work and knit the .rmd file.  
2. Open the .html file and "print" it to a .pdf file in Google Chrome (not Safari).    
3. Go to the class Canvas page and open Gradescope.  
4. Submit your .pdf file to the midterm assignment- be sure to assign the pages to the correct questions.  
5. Commit and push your work to your repository. 