---
title: "Midterm 1 Notes"
output: html_document
date: "2025-02-04"
---

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(naniar)
msleep <- msleep
clean_names(msleep)
```
Before we do anything with our data, we have to load it.
```{r}
#dataexample <- read_csv("/data/dataexample.csv")
```

Summary functions, such as summary() and glimpse() can provide early insights into the data.

```{r}
glimpse(msleep)
```
```{r}
summary(msleep)
```
We can use dplyr functions to manipulate our data and to select items of interest.

Select is used to choose which columns to display.
```{r}
msleep %>% 
  select(name, vore) #only displays the name and vore columns
```
Filter is used to choose certain entries within columns.
```{r}
msleep %>% 
  filter(vore == "carni") #filters all carnivores
```
Within filter, we use == for equal, and != for does not equal.
```{r}
msleep %>%
  filter(vore != "carni") #filters out all carnivores
```
We use & for and. We use | for or. We can use between(column, x, y) to filter values between two chosen numbers.
```{r}
msleep %>% 
  filter(vore == "carni" | vore == "herbi") #filters all carnivores and herbivores
```
```{r}
msleep %>% 
  filter(between(sleep_total, 10, 12)) #filters all animals that sleep between 10 and 12 hours total each day.
```

We can use count() or tabyl() to give a count of how many of each entry there are for a specified column.
```{r}
msleep %>% 
count(vore)
```
```{r}
msleep %>% 
  tabyl(vore)
```
We use summarize() in tandem with group_by() to do many powerful things.
```{r}
msleep %>% 
  summarize(n_vore = n_distinct(vore, na.rm = T)) #tells us how many distinct entries there are for the vore column while also    excluding NAs
```
```{r}
msleep %>% 
  group_by(vore) %>% 
  summarize(average_sleep_vore = mean(sleep_total, na.rm = T), n = n()) #this groups by vore and then calculates the mean sleep total and puts it into a new column called average_sleep_vore. It also displays the total n for each category using n = n().
```
We can even group by multiple variables at once.
```{r}
msleep %>% 
  group_by(vore, conservation) %>% 
  summarize(average_bodywt_vore_conservation = mean(bodywt, na.rm = T), n = n())
```

We can create new variables by manipulating current ones using mutate()
```{r}
msleep %>% 
  select(name, sleep_total) %>% 
  mutate(sleep_total_min = sleep_total * 60) #creates a new variable called sleep_total_min that converts sleep_total from hours   to minutes.
```
We can even select what we want to mutate within a variable using ifelse()
```{r}
msleep %>% 
  select(name, sleep_total) %>% 
  mutate(sleep_designation = ifelse(10 < sleep_total, "lazy bum", "productive member of society"))
```
We can use arrange() to display tables in a specified order.
```{r}
msleep %>% 
  arrange(-sleep_total) #arranges in descending order for sleep_total
```
We can use across() to make more efficient code.
```{r}
msleep %>% 
  summarize(mean_sleep_total = mean(sleep_total, na.rm = T),
            mean_body_wt = mean(bodywt, na.rm = T),
            mean_brain_wt = mean(brainwt, na.rm = T), n = n()) #this is a lot
```
```{r}
msleep %>% 
  summarize(across(c(sleep_total, bodywt, brainwt), \(x) mean(x, na.rm = T)))
```
There are other tools we can use to make this even easier.
```{r}
msleep %>% 
  summarize(across(where(is.numeric), \(x) mean(x, na.rm = T))) #this calculates the mean of all numerics.
```
```{r}
msleep %>% 
  summarize(across(contains("wt"), \(x) mean(x, na.rm = T))) #this calculates the mean of all variables that contain "wt" in them.
```
We can even do multiple calculations at once using list()
```{r}
msleep %>% 
  summarize(across(contains("wt"), list(mean=\(x) mean(x, na.rm = T), sd=\(x) sd(x, na.rm = T)))) #gives us both the mean and standard deviation of variables containing "wt"
```
Sometimes we need to "clean" data to fix issues with NA notation. For this we will load mammal lifehistory data as it has many issues...
```{r}
lifehistories <- read_csv("mammal_lifehistories_v3.csv")
```
```{r}
summary(lifehistories)
```

```{r}
lifehistories %>% 
  summarize(across(everything(), ~ sum(is.na(.))))
```
As you can see, most of the variables have no NAs, which is strange for such a large dataset. Something that is noticeable for the data is that there are many -999 values as minimums for columsn where negative numbers should not be possible, such as "gestation".

One thing we can do is use the naniar package to replace values such as -999 with NA. These two lists are things commonly used to represent NAs.
```{r}
common_na_numbers
common_na_strings
```
We can use these lists with the function replace_with_na_all() to clean up the data.
```{r}
lifehistories <- lifehistories %>% #going back to the original data
  replace_with_na_all(condition = ~.x %in% c(common_na_strings, common_na_numbers)) %>% 
  mutate(newborn=na_if(newborn, "not measured"))
```
Now we can check to see if it worked using miss_var_summary()

```{r}
miss_var_summary(lifehistories)
```
As we can see, things like -999 have been replace with NA.

Another dimension of data tidying is whether our data is long or wide. Wide data is easier to understand and input for humans, but long data is what R usually needs. We will import a dataset in wide format.
```{r}
heartrate <- read_csv("heartrate.csv")
heartrate
```
As you can see, data for drug A, B, C, or D is stored as a column name and not within a column.
We use pivot_longer() to change to a long format where the drug is within a column named "drug".
```{r}
heartrate_long <- heartrate %>% 
  pivot_longer(-patient, #leaves out the patient column
               names_to = "drug", #puts drugs a, b, c, and d into a column called "drug"
               values_to = "heartrate") #puts the values in
heartrate_long
```
We can turn the data back into a wide format with pivot_wider().
```{r}
heartrate_long %>% 
  pivot_wider(names_from = "drug",
              values_from = "heartrate")
```
Sometimes there is more than one variable in a column name. In this case, experiment # and repetition # are combined.
```{r}
qpcr_untidy <- read_csv("qpcr_untidy.csv")
qpcr_untidy
```
We can separate these using names_sep = "separator" within pivot_longer().
```{r}
qpcr_long <- qpcr_untidy %>% 
  pivot_longer(-gene,
               names_to = c("experiment", "replicate"),
               names_sep = "_",
               values_to = "mRNA_expression")
qpcr_long
```
We can also separate at a specific character.
```{r}
qpcr_untidy %>% 
  pivot_longer(-gene,
               names_to = c("experiment", "replicate"),
               names_sep = 5,
               values_to = "mRNA_expression")
```
However the _ at the end of exp is annoying, so we can remove it using str_remove() within mutate().
```{r}
worse <- qpcr_untidy %>% 
  pivot_longer(-gene,
               names_to = c("experiment", "replicate"),
               names_sep = 5,
               values_to = "mRNA_expression") %>% 
  mutate(experiment = str_remove(experiment, "_"))
worse
```
We can also reunite using unite().
```{r}
worse %>% 
  unite(new_column, c(experiment, replicate), sep = "_")
```

We use ggplot for plotting. 
We use geom_point/geom_jitter for scatterplots to show relationships between two continuous variables.
```{r}
lifehistories %>% 
  ggplot(aes(x = gestation, y = `litter size`))+
  geom_point()
```
We use geom_bar for counts, and geom_col to make barplots.
```{r}
lifehistories %>% 
  ggplot(aes(x = order))+
  geom_bar()+
  coord_flip()
```
```{r}
lifehistories %>%
  filter(family == "Bovidae") %>%
  top_n(10, mass) %>% 
  ggplot(aes(x = reorder(species, mass), y = mass))+
  geom_col()+
  coord_flip()
```

We can even color the bars with another variable or by the same variable, and use top_n() to select the top whatever, and use reorder() to make it look nice.
```{r}
lifehistories %>%
  filter(family == "Bovidae") %>%
  top_n(10, mass) %>% 
  ggplot(aes(x = reorder(species, mass), y = mass, fill = Genus))+
  geom_col()+
  coord_flip()
```

We use geom_boxplot to make boxplots to show the ranges that a categorical value takes.
```{r}
lifehistories %>%
  filter(order == "Carnivora") %>% 
  ggplot(aes(x = reorder(family, mass), y = mass))+
  geom_boxplot()+
  scale_y_log10()+
  coord_flip()
```
We should also get rid of scientific notation because it is annoying. We use the following code...
```{r}
options(scipen=999)#cancels scientific notation for the session
```

We can use histograms or density plots to show distributions. We use geom_histogram() or geom_density() to produce these.
```{r}
lifehistories %>% 
  ggplot(aes(x = mass))+
  geom_histogram()+
  scale_x_log10()
```
```{r}
lifehistories %>% 
  ggplot(aes(x = `mass`))+
  geom_density()+
   scale_x_log10()
```
We can even overlay the two.
```{r}
lifehistories %>% 
  ggplot(aes(x = mass))+
  geom_histogram(aes(y = after_stat(density)),color = "black", fill = "wheat", alpha = 0.5)+
  geom_density(color = "purple")+
  scale_x_log10()
```
We can use case_when with mutate to turn a continuous variable into a categorical variable. The bottom is an example because I'm too lazy to adapt it for the specific dataset I've been using.

```{r}
#homerange <- homerange %>% 
#  mutate(mass_category = case_when(log10.mass <= 1.7 ~ "small",
#                                   log10.mass > 1.7 & log10.mass <= 3.33 ~ "median",
#                                   log10.mass > 3.33 ~ "large"))
```

Finally, we can use facet_grid() and facet_wrap() to essentially compare 2 categorical variables to a continuous variable.
```{r}
msleep %>%
  filter(vore == "carni" | vore == "herbi") %>% 
  ggplot(aes(x = order, y = sleep_total, fill = order))+
  geom_boxplot()+
  facet_wrap(~vore, nrow = 2)+ #nrow does rows ncol does columns 
  coord_flip()+
  scale_y_log10()
```
This is a horrible example but yeah.

```{r}
homerange %>% 
  ggplot(aes(x=taxon, y=log10.mass))+
  geom_boxplot()+
  facet_grid(trophic.guild~thermoregulation)+
  theme(axis.text.x = element_text(angle = 60, hjust=1))
```











































