---
title: "Homework 3"
author: "William Holley"
date: "`r Sys.Date()`"
output:
  html_document: 
    theme: spacelab
    keep_md: yes
---

## Instructions
Answer the following questions and/or complete the exercises in RMarkdown. Please embed all of your code and push the final work to your repository. Your report should be organized, clean, and run free from errors. Remember, you must remove the `#` for any included code chunks to run.  

## Load the tidyverse
```{r}
library("tidyverse")
options(scipen=999) # turn off scientific notation
```

## Data
For the homework, we will use data about vertebrate home range sizes. The data are in the class folder and the reference is below.  

**Database of vertebrate home range sizes.**. 
Reference: Tamburello N, Cote IM, Dulvy NK (2015) Energy and the scaling of animal space use. The American Naturalist 186(2):196-211. http://dx.doi.org/10.1086/682070.  
Data: http://datadryad.org/resource/doi:10.5061/dryad.q5j65/1  

**1. Load the data into a new object called `homerange`.**
```{r}
homerange <- read_csv("data/Tamburelloetal_HomeRangeDatabase.csv")
```

**2. What are the dimensions of the dataframe?**  
```{r}
dim(homerange)
```

**3. Are there any NA's in the dataframe? Try using summary to determine which variables have more or less NA's.**
```{r}
summary(homerange)
```
preymass, log10.preymass, and PPMR all have 502 NAs.

**4. What are the names of the columns in the dataframe?**
```{r}
names(homerange)
```

**5. Based on the summary output, do you see anything in the data that looks strange? Think like a biologist and consider the variables.**  

There are many abnormalities and quirks in the data. One variable that caught my eye was the "dimension" variable, because all of its entries are either 2 or 3. I was trying to figure out what the authors of the data meant by "dimension" so I used the DOI number to find their paper and figure it out. "dimension" refers to the foraging dimension of the animal, animals that can forage in 3 dimensions, such as fish or birds, are assigned 3, and animals that forage in "2" dimensions, such as camels or cows, are assigned 2. Although there are no issues with this data itself, I feel like the variable dimension should have a more descriptive name such as "foraging.dimension" to make it more clear. Another issue I noticed is the large amount of NAs within the data. A few are okay, but after a certain point they become detrimental to any calculations performed.

**6. The `min` and `max` functions can be used to find the minimum and maximum values in a vector. Use these functions to find the minimum and maximum values for the variable `mean.mass.g`.**  
```{r}
min(homerange[ ,10])
max(homerange[ ,10])
```

**7. Change the class of the variables `taxon` and `order` to factors and display their levels.**  
```{r}
homerange$taxon <- as.factor(homerange$taxon)
homerange$order <- as.factor(homerange$order)
```

```{r}
levels(homerange$taxon)
```

```{r}
levels(homerange$order)
```

**8. Use `select` to pull out the variables taxon and common.name.**  
```{r}
select(homerange, taxon, order)
```

**9. Use `filter` to pull out all mammals from the data.**
```{r}
filter(homerange, taxon == "mammals")
```

**10. What is the largest `mean.hra.m2` for mammals? This is a very large number! Which animal has this homerange? Look them up online and see if you can figure out why this number is so large.**
```{r}
mammals <- filter(homerange, taxon == "mammals")
max(mammals$mean.hra.m2)
```
```{r}
filter(mammals, 3550830977==mean.hra.m2)
```
The mammal with the largest homerange is the reindeer. This is probably because reindeer live in environments that are sparse, forcing them to travel far distances to find food to eat, as well as to find mates. 


## Knit and Upload
Please knit your work as a .pdf or .html file and upload to Canvas. Homework is due before the start of the next lab. No late work is accepted. Make sure to use the formatting conventions of RMarkdown to make your report neat and clean!  